Online Multiclass Linear Classification with Bandit Feedback

### 2019-01-16 arxiv-version/np-hardness-of-the-weak-labeling-problem.tex
fixing small error in np-hard proof

For example, if we have $S=\{1,2,3\}$, $C=\{c_1, c_2\}$, $c_1 = \{1,2\}$, $c_2=\{2,3\}$, then we create the weak labeling sample set as: 
\begin{align*}
    \{((0,0,0,1),3), ((1,0,0,1),\overline{3}), ((0,1,0,1),\overline{3}), ((0,0,1,1),\overline{3}), ((1,1,0,1),3), ((0,1,1,1),3)\}.
\end{align*}
The following lemma shows that answering this weak labeling problem is equivalent to answering the original set splitting problem. 
\begin{lemma}
Suppose we are given a set splitting problem, and we construct a weak labeling problem as described above. Then the set splitting problem has a solution $\Longleftrightarrow$ the constructed weak labeling problem has a solution. 
\end{lemma}
\begin{proof}
 $(\Longrightarrow)$ Let $S_1, S_2$ be the solution of the set splitting problem. Define
 \begin{align*}
     w_1 = \left(a_1, a_2, \cdots, a_N, -\frac{1}{2}\right), 
 \end{align*}
 where $\forall i\in[N]$, $a_i=1$ if $i\in S_1$ and $a_i=-N$ if $i\notin S_1$. 
 Similarly, define
 \begin{align*}
     w_2 = \left(b_1, b_2, \cdots, b_N, -\frac{1}{2}\right), 
 \end{align*}
 where $\forall i\in[N]$, $b_i=1$ if $i\in S_2$ and $b_i=-N$ if $i\notin S_2$. 
 Finally, define
 \begin{align*}
     w_3 = (0,0,\cdots, 0), 
 \end{align*}
 the zero vector. To see this is a solution for the weak labeling problem, we verify separately for Type 1-3 samples defined above. For Type 1 sample, we have
 \begin{align*}
     \ip{w_3}{x} = 0 > -\frac{1}{2} = \ip{w_1}{x}=\ip{w_2}{x}. 
 \end{align*}
 For a Type 2 sample that corresponds to index $i$, we have either $i\in S_1$ or $i\in S_2$ because $S_1\cup S_2 = [N]$ is guaranteed. Thus, either $a_i=1$ or $b_i=1$. If $a_i=1$ is the case, then
 \begin{align*}
     \ip{w_1}{x}=a_i-\frac{1}{2}=\frac{1}{2} > 0 = \ip{w_3}{x}; 
 \end{align*}
 similarly if $b_i=1$, we have $\ip{w_2}{x}>\ip{w_3}{x}$. \\
 For a Type 3 sample that corresponds to index $j$, Since $c_j \not\subset S_1$, there exists some $i'\in c_j$ and $i'\notin S_1$. Thus we have $x_{i'}=1$, $a_{i'}=-N$, and therefore
 \begin{align*}
     \ip{w_1}{x}= a_{i'}x_{i'} + \sum_{i\in [N]\backslash\{i'\}} a_ix_i - \frac{1}{2} \leq -N + (N-1)-\frac{1}{2} < 0 = \ip{w_3}{x}. 
 \end{align*}
 Because $c_j \not\subset S_2$ also holds, we also have $\ip{w_2}{x}<\ip{w_3}{x}$. This direction is therefore proved. \\
 \ \\
 $(\Longleftarrow)$ Given the solution $w_1, w_2, w_3$ of the weak labeling problem, we define 
 \begin{align*}
     S_1 = \left\{i\in [N] \Big| \ip{w_1-w_3}{(\e_i, 1)} > 0 \right\}, \\
     S_2 = \left\{i\in [N] \Big| \ip{w_2-w_3}{(\e_i, 1)} > 0 \right\}.
 \end{align*}
 It is not hard to see $S_1\cup S_2 = [N]$. Note that $(\e_i, 1)$ is the feature vector for Type 2 samples. Because Type 2 samples all have label $\overline{3}$, for any $i\in [N]$, one of the following must hold: $\ip{w_1-w_3}{(\e_i, 1)}>0$ or $\ip{w_2-w_3}{(\e_i, 1)}>0$. Thus $i\in S_1$ or $i\in S_2$.

 Now we show $\forall j$, $c_j \not\subset S_1$ and $c_j \not\subset S_2$ by contraction. Without loss of generality, suppose there exists some $j$ such that $c_j \subset S_1$. By our definition of $S_1$, we have $\ip{w_1-w_3}{(\e_i, 1)} > 0$ for all $i\in c_j$. Therefore, 
 \begin{align*}
     \sum_{i\in c_j}\ip{w_1-w_3}{\left(\e_i, 1\right)} = \ip{w_1-w_3}{\left(\sum_{i\in c_j}\e_i, |c_j|\right)} > 0. 
 \end{align*}
 Because Type 1 sample has label $3$, we also have
 \begin{align*}
     \ip{w_1-w_3}{\left(\zero, 1\right)} < 0. 
 \end{align*}
 Combining the above two inequalities, we get
 \begin{align*}
     \ip{w_1-w_3}{\left(\sum_{i\in c_j}\e_i, 1\right)} = \ip{w_1-w_3}{\left(\sum_{i\in c_j}\e_i, |c_j|\right)} - (|c_j|-1)\ip{w_1-w_3}{\left(\zero, 1\right)}>0.  
 \end{align*}
 Note that $\left(\sum_{i\in c_j}\e_i, 1\right)$ is a feature vector for Type 3 samples. Thus the above inequality contradicts that Type 3 samples have label 3. Therefore, $c_j \not\subset S_1$. Similarly we can prove $c_j \not\subset S_2$. 
\end{proof}

>>>>

For example, if we have $S=\{1,2,3\}$, $C=\{c_1, c_2\}$, $c_1 = \{1,2\}$, $c_2=\{2,3\}$, then we create the weak labeling sample set as: 
\begin{align*}
    \{((0,0,0,1),3), ((1,0,0,1),\overline{3}), ((0,1,0,1),\overline{3}), ((0,0,1,1),\overline{3}), ((1,1,0,1),3), ((0,1,1,1),3)\}.
\end{align*}
The following lemma shows that answering this weak labeling problem is equivalent to answering the original set splitting problem. 
\begin{lemma}
Suppose we are given a set splitting problem, and we construct a weak labeling problem as described above. Then the set splitting problem has a solution $\Longleftrightarrow$ the constructed weak labeling problem has a solution. 
\end{lemma}
\begin{proof}
 $(\Longrightarrow)$ Let $S_1, S_2$ be the solution of the set splitting problem. Define
 \begin{align*}
     w_1 = \left(a_1, a_2, \cdots, a_N, -\frac{1}{2}\right), 
 \end{align*}
 where $\forall i\in[N]$, $a_i=1$ if $i\in S_1$ and $a_i=-N$ if $i\notin S_1$. 
 Similarly, define
 \begin{align*}
     w_2 = \left(b_1, b_2, \cdots, b_N, -\frac{1}{2}\right), 
 \end{align*}
 where $\forall i\in[N]$, $b_i=1$ if $i\in S_2$ and $b_i=-N$ if $i\notin S_2$. 
 Finally, define
 \begin{align*}
     w_3 = (0,0,\cdots, 0), 
 \end{align*}
 the zero vector. To see this is a solution for the weak labeling problem, we verify separately for Type 1-3 samples defined above. For Type 1 sample, we have
 \begin{align*}
     \ip{w_3}{x} = 0 > -\frac{1}{2} = \ip{w_1}{x}=\ip{w_2}{x}. 
 \end{align*}
 For a Type 2 sample that corresponds to index $i$, we have either $i\in S_1$ or $i\in S_2$ because $S_1\cup S_2 = [N]$ is guaranteed. Thus, either $a_i=1$ or $b_i=1$. If $a_i=1$ is the case, then
 \begin{align*}
     \ip{w_1}{x}=a_i-\frac{1}{2}=\frac{1}{2} > 0 = \ip{w_3}{x}; 
 \end{align*}
 similarly if $b_i=1$, we have $\ip{w_2}{x}>\ip{w_3}{x}$. \\
 For a Type 3 sample that corresponds to index $j$, Since $c_j \not\subset S_1$, there exists some $i'\in c_j$ and $i'\notin S_1$. Thus we have $x_{i'}=1$, $a_{i'}=-N$, and therefore
 \begin{align*}
     \ip{w_1}{x}= a_{i'}x_{i'} + \sum_{i\in [N]\backslash\{i'\}} a_ix_i - \frac{1}{2} \leq -N + (N-1)-\frac{1}{2} < 0 = \ip{w_3}{x}. 
 \end{align*}
 Because $c_j \not\subset S_2$ also holds, we also have $\ip{w_2}{x}<\ip{w_3}{x}$. This direction is therefore proved. \\
 \ \\
 $(\Longleftarrow)$ Given the solution $w_1, w_2, w_3$ of the weak labeling problem, we define 
 \begin{align*}
     S_1 &= \left\{i\in [N] \Big| \ip{w_1-w_3}{(\e_i, 1)} > 0 \right\}, \\
     S_2 &= \left\{i\in [N] \Big| \ip{w_2-w_3}{(\e_i, 1)} > 0 \text{\ and\ } i\notin S_1 \right\}.
 \end{align*}
 It is not hard to see $S_1 \cap S_2 = \emptyset$ and $S_1\cup S_2 = [N]$. The former is because $S_2$ only includes elements that are not in $S_1$. For the latter, note that $(\e_i, 1)$ is the feature vector for Type 2 samples. Because Type 2 samples all have label $\overline{3}$, for any $i\in [N]$, one of the following must hold: $\ip{w_1-w_3}{(\e_i, 1)}>0$ or $\ip{w_2-w_3}{(\e_i, 1)}>0$. This implies $i\in S_1$ or $i\in S_2$.

 Now we show $\forall j$, $c_j \not\subset S_1$ and $c_j \not\subset S_2$ by contraction. Assume there exists some $j$ such that $c_j \subset S_1$. By our definition of $S_1$, we have $\ip{w_1-w_3}{(\e_i, 1)} > 0$ for all $i\in c_j$. Therefore, 
 \begin{align*}
     \sum_{i\in c_j}\ip{w_1-w_3}{\left(\e_i, 1\right)} = \ip{w_1-w_3}{\left(\sum_{i\in c_j}\e_i, |c_j|\right)} > 0. 
 \end{align*}
 Because Type 1 sample has label $3$, we also have
 \begin{align*}
     \ip{w_1-w_3}{\left(\zero, 1\right)} < 0. 
 \end{align*}
 Combining the above two inequalities, we get
 \begin{align*}
     \ip{w_1-w_3}{\left(\sum_{i\in c_j}\e_i, 1\right)} = \ip{w_1-w_3}{\left(\sum_{i\in c_j}\e_i, |c_j|\right)} - (|c_j|-1)\ip{w_1-w_3}{\left(\zero, 1\right)}>0.  
 \end{align*}
 Note that $\left(\sum_{i\in c_j}\e_i, 1\right)$ is a feature vector for Type 3 samples. Thus the above inequality contradicts that Type 3 samples have label 3. Therefore, $c_j \subset S_1$. If we assume there exists some $c_j \not\subset S_2$, same arguments apply and also lead to contradiction. 
\end{proof}
